Two Decades of Smalltalk VM Development

### 2018-08-16 paper.tex
pushing latest comments on section 2 for today

When programming with Smalltalk, the programmer usually starts from a snapshot which contains the core libraries, the development environment and the application under development. Development of the application consists essentially in writing and editing code, which effectively installs, modifies and removes classes and compiled methods from the set of live objects. Programming may be done live, as the application under development is running, for example creating or revising methods in the debugger, and continuing.  Objects may have their shape changed on the fly as instance variables are added and removed.  When the programmer chooses, the developer takes a new snapshot, which includes its changes, and may include application instances under development [MAYBE THIS SHOULD BE DESCRIBED BELOW IN SNAPSHOTS]. Deployment is performed using a snapshot containing the deployed application and the required core libraries (unused libraries and development tools may be removed from the deployed snapshot to decrease memory footprint and for security purposes).  \footnote{Code is organised in packages which are saved to an external version control system, in our case Monticello citation?  wiresong.ca}

@@

Figure \ref{fig:SimulatedMemory} describes, from a high-level perspective, the memory used by \openSmalltalkVM at runtime and the simulated counter-part. Let's detail briefly the memory used by the production VM, on the top of the figure. On the left side, which is usually the low addresses, we can find Text, the memory section holding the native code of the VM (the interpreter, the memory manager and the JIT itself, but not the code compiled by the JIT). Then, Data holds initialized and uninitialized data, mainly the C variables used globally in the VM. On higher addresses, we can find the beginning of the memory managed by the VM. At start-up, the VM mmaps a memory region used for the executable code generated by the JIT (optional executable section, exists only if the JIT is enabled), for the young objects and for the old objects present in the snapshot started plus a little more space for the first tenures\footnote{A tenure is the process of promoting young objects to old objects in a generational GC.} to be performed without triggering the full garbage collector. Later during execution, as old space grows, new memory regions are mmap'ed on higher addresses to store other old objects. Usually at very high addresses, we can find the C stack. Room in the C stack is alloca'ed at VM start-up to hold the Smalltalk stack. Both stacks are disjointed and managed differently.

In the simulator, the heap is simulated as a large contiguous byte array. References between objects are effectively indexes inside the byte array instead of pointers. All the Slang variables, normally translated to C variabled, are simulated as Smalltalk objects. They used specific classes, such as \emph{CArrayAccessor} over normal Smalltalk classes, to emulate the C behavior (only array accesses are available in C, not high level iterator APIs, etc.). The Slang code is executed as Smalltalk code. The Smalltalk stack is represented in \openSmalltalkVM as a double linked list of stack pages which are maintained by the VM. Each stack page is represented as a Smalltalk object in the simulator.

All the Slang code is implemented in multiple Smalltalk classes, to organise the code and to add flexibility through polymorphism. For example, the AbstractCompactor class has two subclasses, one implements a sweep algorithm and the other a compact one. For production, at Slang-to-C compilation time, all the code is compiled in a single C file. All the flexibility is removed, using the same example, the VM developer chooses at this moment if he wants to compile a VM with a sweep or compact algorithm. No polymorphism is available at runtime. However, since polymorphism is available in the simulator, it can be abused for debugging purposes. Still re-using the same example, the AbstractCompactor class has also simulation specific subclasses. Such versions typically express additional constraints in the form of assertions which can be written in plain Smalltalk without restrictions to easily express complex constraints. They also keep specific values live so they can be accessed at debugging time.

>>>>

When programming with Smalltalk, the programmer usually starts from a snapshot which contains the core libraries, the development environment and the application under development. Development of the application consists essentially in writing and editing code, which effectively installs, modifies and removes classes and compiled methods from the set of live objects. Programming may be done live, as the application under development is running, for example creating or revising methods in the debugger, and continuing.  Objects may have their shape changed on the fly as instance variables are added and removed.
\egb{there is overlap here with respect to the previous paragraph. We already knew that a snapshot can be made during a coding session, and what a snapshot includes}
  When the programmer chooses, he can take a new snapshot, which includes his code changes, and may include application instances under development [MAYBE THIS SHOULD BE DESCRIBED BELOW IN SNAPSHOTS]. \egb{deployment here refers to starting a Smalltalk system from the previous paragraph, right?}Deployment is performed using a snapshot containing the deployed application and the required core libraries (unused libraries and development tools may be removed from the deployed snapshot to decrease memory footprint and for security purposes).  \footnote{Code is organised in packages which are saved to an external version control system, in our case Monticello citation?  wiresong.ca}

@@

Figure \ref{fig:SimulatedMemory} describes, from a high-level perspective, the memory used by \openSmalltalkVM at runtime and the simulated counter-part. Let us detail briefly the memory used by the production VM, on the top of the figure. On the left side, which is usually the low addresses, we can first find Text, the memory section holding the native code of the VM (the interpreter, the memory manager and the JIT itself, but not the code compiled by the JIT). Then, Data holds initialized and uninitialized data, mainly the C variables used globally in the VM. On higher addresses, we can find the beginning of the memory managed by the VM.

At start-up, the VM mmaps a memory region used for the executable code generated by the JIT (optional executable section, exists only if the JIT is enabled), for the young objects and for the old objects present in the snapshot started plus a little more space for the first \emph{tenures}\footnote{A tenure is the process of promoting young objects to old objects in a generational GC.} to be performed without triggering the full garbage collector. Later during execution, as old space grows, new memory regions are mmap'ed on higher addresses to store other old objects. Usually at very high addresses, we can find the C stack. Room in the C stack is allocated at VM start-up to hold the Smalltalk stack. Both stacks are disjointed and managed differently.

In the simulator, the heap is simulated as a large contiguous byte array. References between objects are effectively indexes inside the byte array instead of pointers. All the Slang variables, normally translated to C variables, are simulated as Smalltalk objects. They used specific classes, such as \emph{CArrayAccessor} over normal Smalltalk classes, to emulate the C behavior (only array accesses are available in C, not high level iterator APIs, etc.). The Slang code is executed as Smalltalk code. The Smalltalk stack is represented in \openSmalltalkVM as a double linked list of stack pages which are maintained by the VM. Each stack page is represented as a Smalltalk object in the simulator.

All the Slang code is implemented in multiple Smalltalk classes, to organise the code and to add flexibility through polymorphism. For example, the \texttt{AbstractCompactor} class has two subclasses, one implementing a sweep algorithm and the other a compact one. For production, at Slang-to-C compilation time, all the code is compiled in a single C file. All the flexibility is removed, using the same example, the VM developer chooses at this moment if he wants to compile a VM with a sweep or compact algorithm. No polymorphism is available at runtime. However, since polymorphism is available in the simulator, it can be reused for debugging purposes. Still in our example, the \texttt{AbstractCompactor} class has also simulation specific subclasses. Such versions typically express additional constraints in the form of assertions which can be written in plain Smalltalk without restrictions to easily express complex constraints. They also keep specific values live so they can be accessed at debugging time.
